The ``Dark Side'' of Phylogenetic Comparative Methods

### 2014-08-21 Manuscript/dark.tex
Reordered introduction

\documentclass[a4paper,12pt]{article}
\usepackage[osf]{mathpazo} % palatino
\usepackage{ms}     % load the template
\usepackage[round]{natbib} % author-year citations
\usepackage{graphicx}
\pagenumbering{arabic}

\title{The ``Dark Side'' of Phylogenetic Comparative Methods}
\author{
Richard G. FitzJohn$^{1}$, Matthew W. Pennell$^{2}$, others,\\ and Natalie Cooper$^{3,4,*}$
}
\date{}
\affiliation{\noindent{\footnotesize
$^1$ Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia \\
$^2$ Institute for Bioinformatics and Evolutionary Studies, University
of Idaho, Moscow, ID 83844, U.S.A.\\
$^3$ School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\ 
$^4$ Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland.\\
$^*$ Corresponding author: ncooper@tcd.ie; Zoology Building, Trinity College Dublin, Dublin 2, Ireland. Fax: +353 1 677 8094; Tel: +353 1 896 1926.\\
}}

\runninghead{The "dark side" of PCMs} % "Point of view" for Syst Biol
\keywords{PCM, assumption, etc.}

%%%%%
% From Syst Biol template
\linespread{1.66}
\raggedright
\setlength{\parindent}{0.5in}

@@

\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}

\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}

@@

%%%%%

@@

\mstitlepage
\parindent=1.5em
\addtolength{\parskip}{.3em}

@@

Phylogenetic comparative methods (PCMs) were initially developed in the 1980s to deal with the statistical non-independence of species in comparative analyses (e.g. \citealp{felsenstein1985phylogenies,grafen1989phylogenetic}). Since then PCMs have been extended to investigate evolutionary pattern and process, and include methods for investigating drivers of diversification, the tempo and mode of trait evolution, and models of speciation and extinction (see reviews in \citealp{o2012evolutionary, pennell2013integrative}). PCMs have become extremely popular over recent years; in 2013 alone Harvey and Pagel's \citeyearpar{harvey1991comparative} book on \textit{The Comparative Method in Evolutionary Biology} was cited 195 times (Google Scholar, 22nd May 2014), and at the 2014 Evolution meeting in Raleigh NC, we estimate that over XX %need to count these!
talks used PCMs of one form or another.

PCMs also have a ``dark side''; they make various assumptions and suffer from biases in the same way as any statistical method. Unfortunately, the more popular these methods become, the less awareness methods-users seem to have of this ``dark side''. Increasingly assumptions and biases are inadequately assessed in empirical studies, leading to poor model fits and misinterpreted results. Clearly more effort is needed to bridge the widening gap between methods-users and methods-developers, but who should take responsibility for such actions? And why has this gap developed in the first place? Here we explore these issues and make suggestions for improvements.\\ %needs some rewording

\subsection{Why aren't more methods-users aware of the ``dark side'' of PCMs?}

@@

\begin{enumerate}
\item Assumptions and biases are not mentioned in the literature.
\item Assumptions and biases are mentioned in the literature, but the literature is too technical for the novice user and/or pertinent details are hidden within the text.
\item Methods users have bypassed the literature gone straight to the implementation of the method.
\end{enumerate}

Note that we assume that methods-users ignore assumptions and biases because they are unaware of them. Of course, there may be methods-users who are perfectly aware of these issues but choose to sweep them under the carpet to get their work published. Although we believe there is no excuse for doing this, we also recognize that the ``publish or perish'' culture of academia may pressurize nontenured scientists into this. Hopefully some of the changes we suggest below will also help with this problem.

\subsubsection{1. Assumptions and biases are not mentioned in the literature}

@@

\subsubsection{2. Literature is too technical or dense}

@@

\subsubsection{3. Users jump straight to the implementation of the method}

@@

@@

\section{Funding}
This work was supported by The European Commission CORDIS Seventh Framework Program (FP7) Marie Curie CIG grant, proposal number: 321696 (NC)

\section{Acknowledgments}
Thanks to Star Wars for the ``hilarious'' jokes.

>>>>

% Preamble
\documentclass[a4paper,12pt]{article}

\usepackage[osf]{mathpazo} % palatino
\usepackage{ms}            % load the template
\usepackage[round]{natbib} % author-year citations
\usepackage{graphicx}      
\pagenumbering{arabic}

% From Syst Biol template
\linespread{1.66}
\raggedright
\setlength{\parindent}{0.5in}

@@

\renewcommand{\section}[1]{
  \bigskip
  \begin{center}
  \begin{Large}
  \normalfont\scshape #1
  \medskip
  \end{Large}
  \end{center}
}

\renewcommand{\subsection}[1]{
  \bigskip
  \begin{center}
  \begin{large}
  \normalfont\itshape #1
  \end{large}
  \end{center}
}

@@

% Title page information

\title{The ``Dark Side'' of Phylogenetic Comparative Methods}
\author{
  Richard G. FitzJohn$^{1}$, Cecile An\'{e}$^{2}$, Wayne Maddison$^{3}$,\\ Matthew W. Pennell$^{4}$, Gavin H. Thomas$^{5}$, April Wright$^{6}$,\\ and Natalie Cooper$^{7,8,*}$
}
\date{}
\affiliation{\noindent{\footnotesize
$^1$ Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia. \\
$^2$ Wisconsin\\
$^3$ UBC\\
$^4$ Institute for Bioinformatics and Evolutionary Studies, University
of Idaho, Moscow, ID 83844, U.S.A.\\
$^5$ Sheffield\\
$^6$ Austin\\
$^7$ School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\ 
$^8$ Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland.\\
$^*$ Corresponding author: ncooper@tcd.ie; Zoology Building, Trinity College Dublin, Dublin 2, Ireland. Fax: +353 1 677 8094; Tel: +353 1 896 1926.\\
}}

\runninghead{The ``Dark Side'' of PCMs}
\keywords{PCM, assumption, etc.}

% End of preamble

@@

\mstitlepage
\parindent = 1.5em
\addtolength{\parskip}{.3em}

@@

A long time ago in a galaxy far, far away - also known as the 1980s - phylogenetic comparative methods (PCMs) were developed to deal with the statistical non-independence of species in comparative analyses (e.g. \citealp{felsenstein1985phylogenies,grafen1989phylogenetic}). 
Since then PCMs have been extended to investigate evolutionary pattern and process, and include methods for investigating drivers of diversification, the tempo and mode of trait evolution, and models of speciation and extinction (see reviews in \citealp{o2012evolutionary, pennell2013integrative}). 
PCMs have become extremely popular over recent years; the number of papers featured the phrase `phylogenetic comparative' has increased exponentially since the 1980s (Figure \ref{PCMCitations}). 
With new methods being published almost weekly, there has never been a better time to be a comparative biologist.

\begin{figure}[h]
\centering
\includegraphics[width = 8cm]{PCMCitations.png}
\caption{PCM papers through time}
\label{PCMCitations}
\end{figure}

Unfortunately, PCMs also have a ``dark side''; they make various assumptions and suffer from biases in exactly the same way as any statistical method \citep{freckleton2009seven,boettiger2012your}. 
The more popular these methods become, the less awareness methods-users seem to have of this dark side. Increasingly assumptions and biases are inadequately assessed in empirical studies, leading to poor model fits and misinterpreted results. 
Additionally, little consideration is given to whether using a PCM is really appropriate for the question at hand \citep{losos2011seeing}. 
To paraphrase \citet{blomberg2012independent}, the enthusiastic application of PCMs has greatly outpaced our theoretical understanding of the methods and their relations to the rest of statistical theory.

% Do we need more explicit comments showing we're not the only people to think this, nor the first to say it?

One major issue is that two groups of people are involved with PCMs. 
At one extreme, methods-developers (wookies) are involved with developing new methods and implementing them. They are aware of the limitations of their methods, and the assumptions that underly them. 
% Bit of a stretch but we can pretend right? :)
At the other extreme, naive methods-users (ewoks) have no knowledge of these complexities, they just want to use PCMs on their data. 
They may be graduate students or more established scientists who have been asked to apply a phylogenetic correction by a reviewer. 
Of course the majority of PCM users fall somewhere between these two extremes on the ``wookie-ewok'' spectrum; many methods-developers are developing methods with a specific empirical question in mind, while there are many advanced methods-users who have detailed knowledge of the limitations of the methods they use. 
Although methods-developers (and advanced methods-users) are aware of the problems with PCMs, the tools and approaches used to fit models are often far more user-friendly and better documented than the methods used to to assess whether that model fit is reasonable. 
Clearly more effort is needed to bridge the widening gap between methods-users and methods-developers, but who should take responsibility for such actions? And why has this gap developed in the first place? Here we explore these issues and make suggestions for improvements.\\ %needs some rewording

\section{Why aren't more methods-users aware of the ``dark side'' of PCMs?}

@@

\begin{enumerate}
\item Not everything is mentioned in the literature.
\item The literature is too technical for the novice user and/or pertinent details are hidden within the text.
\item Methods users have bypassed the literature gone straight to the implementation of the method.
\end{enumerate}

%Note that we assume that methods-users ignore assumptions and biases because they are unaware of them. Of course, there may be methods-users who are perfectly aware of these issues but choose to sweep them under the carpet to get their work published. Although we believe there is no excuse for doing this, we also recognize that the ``publish or perish'' culture of academia may pressurize nontenured scientists into this. Hopefully some of the changes we suggest below will also help with this problem.

\subsection{1. Not everything is mentioned in the literature}

Program defaults are another thing which become folklore although tey are often arbitrary.

@@

\subsubsection{2. Literature is technical, dense and numerous}

@@

\subsubsection{3. Important information is hard to locate}

\subsubsection{4. Users jump straight to the implementation of the method}

@@

\subsection{An example using contrasts}
Probably the most well-known, PCM is Felsenstein's \citeyearpar{felsenstein1985phylogenies} phylogenetically independent contrasts (PIC). shortly afterwards Grafen introduced the phylogenetic regression also known as phylogenetic generalised least squares (PGLS). Later papers confirmed that these methods are actually equivalent if a model of Brownian motion evolution is assumed (although the mathematical proof for this is "awkward"; \citet{rohlf2001comparative,blomberg2012independent})- with PIC being a special case of PGLS \citet{blomberg2012independent,rohle2006comment,rohlf2001comparative,garland2000using}.

Many papers have discussed the assumptions of this method at length.\citealp{diaz1996testing} list these assumptions as 1) topology is perfect; 2) branch lengths are perfect and 3) the model of evolution is Brownian. Felsenstein points out the assumption of Brownian motion may not be very useful in his original paper \citet[p.13]{felsenstein1985phylogenies}. This was followed up by numerous publications in the 1990s discussing this assumption and providing multiple tests to see if the assumption was correct. Many of these papers are extremely explicit about the need to investigate these assumptions.

Assumptions of tree accuracy are less commonly discussed, partly because evidence suggests they make very little difference. Polytomies are dealt with though.

@@

%A New Hope joke...

\section{Funding}
This work was supported by The European Commission CORDIS Seventh Framework Program (FP7) Marie Curie CIG grant, proposal number: 321696 (NC).

\section{Acknowledgments}
Thanks to the Society of Systematic Biologists for funding our symposium at Evolution 2014. We also thank George Lucas for the ``hilarious'' Star Wars jokes.
